---
title: "Final project"
author: "Charlene Isimbi Kanyange"
date: "2025-05-30"
output: html_document
---

1. List of issues in this dataset:

. Some columns contain missing values.
. Inconsistent coding is present across records, such as variations like D&V, D;V, or D/V.
. Different cases are sometimes assigned the same identification number.
. Date formats vary across records.
. The age variable includes both numbers and text, even though it should only contain numeric values.
. In the second location, certain variables were not recorded for all cases, resulting in missing data.


```{r}
library(readxl)
Cholera_ds <- read_excel("C:/Users/korte/Desktop/Chrln/Auca/R/DataFile.xls")
str(Cholera_ds)
summary(Cholera_ds)
```

```{r}

Cholera_ds$`Case No.` <- 1:2066
```


2. Harmonize the variabe "Sex"
```{r}
Cholera_ds$Sex <- tolower(Cholera_ds$Sex)
Cholera_ds$Sex <- ifelse(Cholera_ds$Sex %in% c("m", "male"), "Male",
                   ifelse(Cholera_ds$Sex %in% c("f", "female"), "Female", NA))
Cholera_ds
```

3. Grouping age
```{r}
Cholera_ds$Age <- tolower(Cholera_ds$Age)
View(Cholera_ds)
```

```{r}
# Extract values
library(stringr)
Cholera_ds$Years <- as.numeric(str_extract(Cholera_ds$Age, "\\d*\\.?\\d+(?=\\s*(yrs|year|yr|y))"))
Cholera_ds$Months <- as.numeric(str_extract(Cholera_ds$Age, "\\d*\\.?\\d+(?=\\s*(months|mo|m|mths|M|ths))"))
Cholera_ds$Plain <- as.numeric(str_extract(Cholera_ds$Age, "^\\d*\\.?\\d+$"))
```

```{r}
# Use plain numbers as years if other values are missing
Cholera_ds$Years <- ifelse(is.na(Cholera_ds$Years), Cholera_ds$Plain, Cholera_ds$Years)

# Compute total months and divide by 12 to get age in years
library(dplyr)
Cholera_ds$Years[is.na(Cholera_ds$Years)] <- 0
Cholera_ds$Months[is.na(Cholera_ds$Months)] <- 0
Cholera_ds$Plain[is.na(Cholera_ds$Plain)] <- 0
Cholera_ds$Age <- round((Cholera_ds$Years * 12 + Cholera_ds$Months)/12,2)
Cholera_ds <- Cholera_ds %>% select(-Years, -Months, -Plain)
Cholera_ds
```

```{r}
#Grouping
Cholera_ds$Age_group <- ifelse(Cholera_ds$Age < 5, "Less than 5 years",
                  ifelse(Cholera_ds$Age < 15, "5-14 years",
                  ifelse(Cholera_ds$Age < 25, "15-24 years",
                  ifelse(Cholera_ds$Age < 50, "25-49 years",
                  ifelse(Cholera_ds$Age >= 50, "50 years and above","Not Given")))))
Cholera_ds
```


4.Produce outputs for briefing of the Minister based on his interest
```{r}
#Total number of cases by province
Cholera_ds %>% count(Province, District) %>% arrange(desc(n))
```

```{r}
#Number of death and the case fatality rate
```

```{r}
#The most and less affected age group
library(ggplot2)
age_summary <- group_by(Cholera_ds, Age_group) %>%
  count(Age_group, sort = TRUE)
age_summary
ggplot(age_summary, aes(x = Age_group, y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), vjust = -0.5) +
  labs(title = "Cases by Age Group",
       x = "Age Group", y = "Number of Cases") +
  theme_minimal()
```

```{r}
#Gender distribution
gender_summary <- Cholera_ds %>% count(Sex)
ggplot(gender_summary, aes(x = Sex, y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), vjust = -0.2) +
  labs(title = "Cases by Gender",
       x = "Gender", y = "Number of Cases") +
  theme_minimal()
```

```{r}
#Analysis of cases in different groups across different age groups,genders and districts

chisq.test(table(Cholera_ds$Age_group))
chisq.test(table(Cholera_ds$Sex))
chisq.test(table(Cholera_ds$District))
```

```{r}
#Epidemic curve
# Adjust if format differs
library(lubridate)
Cholera_ds$OnsetDate <- dmy(Cholera_ds$`Date of onset`) 
```

```{r}
# Convert all to character (in case some are numbers)
Cholera_ds$`Date of onset` <- as.character(Cholera_ds$`Date of onset`)

# Identify numeric Excel-style dates
is_numeric_date <- grepl("^\\d+\\.?\\d*$", Cholera_ds$`Date of onset`)

# Create a new parsed date column
Cholera_ds$ParsedDate <- NA

# Convert Excel numeric dates
Cholera_ds$ParsedDate[is_numeric_date] <- as.Date(as.numeric(Cholera_ds$`Date of onset`[is_numeric_date]), origin = "1899-12-30")

# Convert normal date strings using dmy()
Cholera_ds$ParsedDate[!is_numeric_date] <- dmy(Cholera_ds$`Date of onset`[!is_numeric_date])
 Cholera_ds
```

```{r}
# Fix just those that failed earlier
is_failed <- is.na(Cholera_ds$ParsedDate) & grepl("^\\d{1,2}/\\d{1,2}/\\d{2}$", Cholera_ds$`Date of onset`)
Cholera_ds$ParsedDate[is_failed] <- mdy(Cholera_ds$`Date of onset`[is_failed])
Cholera_ds
```

```{r}
# Ensure date is in Date format
Cholera_ds$ParsedDate <- as.Date(Cholera_ds$ParsedDate)
# Create month and year columns
Cholera_ds <- Cholera_ds %>%
  mutate(Month = floor_date(ParsedDate, "month"),
         Year = year(ParsedDate))
monthly_cases <- Cholera_ds %>%
  count(Month)
yearly_cases <- Cholera_ds %>%
  count(Year)


monthly_cases <- monthly_cases %>%
  mutate(Group = "Monthly")

yearly_cases_line <- yearly_cases %>%
  mutate(Month = as.Date(paste0(Year, "-01-01")),
         Group = "Yearly")

# Combine both into one dataframe
combined <- bind_rows(monthly_cases, yearly_cases_line)

# Plot both trends in one graph
ggplot(combined, aes(x = Month, y = n, color = Group)) +
  geom_line(size = 1.2) +
  labs(title = "Monthly and Yearly Trends of Cholera Cases",
       x = "Date", y = "Number of Cases", color = "Trend Level") +
  scale_color_manual(values = c("Monthly" = "steelblue", "Yearly" = "lightgreen")) +
  theme_minimal()
```

```{r}
#Distribution by outcome and the recovery rate
Cholera_ds$Outcome<-tolower(Cholera_ds$Outcome)
table(Cholera_ds$Outcome)
recovery_rate <- round(nrow(filter(Cholera_ds, Cholera_ds$Outcome == "discharged")) / total_cases * 100,2)
recovery_rate
```

5. The necessary figures that allows visualization of the outbreak situation
```{r}

```

6. Risk factors driving the spread of the outbreak?
```{r}
Cholera_ds$`Water treatment`<- tolower(Cholera_ds$`Water treatment`)
Cholera_ds$`Water treatment`[Cholera_ds$`Water treatment` %in% c("boil", "boiling")] <- "Boiling"
```

```{r}
Cholera_ds$`Water source`<-tolower(Cholera_ds$`Water source`)
```

```{r}
model <- glm(Outcome == "died" ~ Age + Sex  + `Water source` + `Water treatment`, data = Cholera_ds, family = "binomial")
summary(model)
```
Age is the only statistically significant predictor:

Estimate = 0.04018 means that for every one-year increase in age, the log-odds of death from cholera increase by 0.04.

This means that each additional year of age is associated with a 4.1% increase in the odds of dying from cholera.

Sex, water source, and water treatment were not significant predictors in this model.

The model suffers from extensive missing data, reducing power and reliability.

Caution is advised in interpreting results—especially estimates with large standard errors.


7. Recommendations:

Direct hygiene promotion efforts, access to safe water, and mobile healthcare services toward individuals aged 25–49.

Prioritize District 2 with initiatives such as clean water provision, sanitation improvements, hygiene awareness programs, mobile medical units, and distribution of hygiene kits.

Strengthen case reporting by providing better training for data collectors to improve the effectiveness of outbreak response.

Allocate resources to develop water and sanitation systems, addressing root causes and reducing the risk of future outbreaks.
